---
title: "Troubleshooting Sensitivity Analysis"
output: html_document
date: '2023-05-20'
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r prereq, echo = FALSE}
library(ggplot2)
library(brms)

set.seed(19)
```
### Background on Sensitivity Analysis

I've performed the sensitivity analysis for a simplified two plant species + one microbial taxon system by randomly generating parameter values for each of the microbe-independent competition coefficients ($c$), the microbial cultivation rates ($\phi$), and the microbial effects on plants ($\sigma$). I used the following distributions for each:

$$
c \sim U(-2, 0)
$$
$$
\phi \sim U(0, 20)
$$
$$
\sigma \sim U(-.1, .1)
$$

I generated 100,000 random values from these distributions, and I calculated the overall competition coefficients as $\alpha_{ij} = c_{ij} + \sigma_i*\phi_j$. I also calculated the niche differences as $1 - \sqrt{\frac{\alpha_{ij}\alpha{ji}}{\alpha_{ii}\alpha_{jj}}}$ and the niche differences as the larger of two competive ratios ($\sqrt{\frac{\alpha_{ij}\alpha{ii}}{\alpha_{ji}\alpha{jj}}}$) which corresponds to the fitness ratio of the superior competitor over the inferior competitor. Finally, I predicted whether the species would coexist or not based on the relationships between the niche and fitness differences. I ran a linear regression with gaussian error distributions for estimating the niche differences and fitness inequalities ($|FI - 1|$) as a function of the scaled parameters and an additional linear regression with bournoulli error distributions for predicting whether the species would coexist as a function of the scaled parameters. I then extracted the median and upper + lower 95% CIs for the posterior distributions of each of the coefficients, which represent the effect size for each of these parameters on the response variable, and compared them in the following plot:

```{r echo = FALSE}
# setting the range for the parameter distributions

c_min <- -2
c_max <- 0
sig_min <- -.1
sig_max <- .1
phi_min <- 0
phi_max <- 20

# some empty vectors to throw values into
cAAs <- c()
cBAs <- c()
cBBs <- c()
cABs <- c()
sigAs <- c()
sigBs <- c()
phiAs <- c()
phiBs <- c()
alphaAAs <- c()
alphaBAs <- c()
alphaBBs <- c()
alphaABs <- c()
NDs <- c()
FIs <- c()

# generating data from 5000 runs
runs <- 100000
for(i in 1:runs){
  cAAs[i] <- runif(1, min = c_min, max = c_max)
  cBAs[i] <- runif(1, min = c_min, max = c_max)
  cBBs[i] <- runif(1, min = c_min, max = c_max)
  cABs[i] <- runif(1, min = c_min, max = c_max)
  
  sigAs[i] <- runif(1, min = sig_min, max = sig_max)
  sigBs[i] <- runif(1, min = sig_min, max = sig_max)
  
  phiAs[i] <- runif(1, min = phi_min, max = phi_max)
  phiBs[i] <- runif(1, min = phi_min, max = phi_max)
  
  alphaAAs[i] <- cAAs[i]+sigAs[i]*phiAs[i]
  alphaBAs[i] <- cBAs[i]+sigBs[i]*phiAs[i]
  alphaBBs[i] <- cBBs[i]+sigBs[i]*phiBs[i]
  alphaABs[i] <- cABs[i]+sigAs[i]*phiBs[i]
  
  NDs[i] <- 1 - sqrt((alphaABs[i]*alphaBAs[i])/(alphaAAs[i]*alphaBBs[i]))
  FIs[i] <- ifelse(sqrt((alphaAAs[i]*alphaABs[i])/(alphaBBs[i]*alphaBAs[i])) > sqrt((alphaBBs[i]*alphaBAs[i])/(alphaAAs[i]*alphaABs[i])),
                   sqrt((alphaAAs[i]*alphaABs[i])/(alphaBBs[i]*alphaBAs[i])), sqrt((alphaBBs[i]*alphaBAs[i])/(alphaAAs[i]*alphaABs[i])))
  
}

dat <- cbind.data.frame(cAA = scale(abs(cAAs)), cBA = scale(abs(cBAs)), 
                        cBB = scale(abs(cBBs)), cAB = scale(abs(cABs)),
                        sigA = scale(sigAs), sigB = scale(sigBs), 
                        phiA = scale(phiAs), phiB = scale(phiBs), 
                        alphaAA = alphaAAs, alphaBA = alphaBAs, 
                        alphaBB = alphaBBs, alphaAB = alphaABs, 
                        ND = NDs, FI = FIs)

# adding a binary coexistence variable for logistic regression
dat$coexist <- ifelse(dat$FI > -dat$ND + 1 & dat$FI < 1/(-dat$ND+1), 1, 0)

# running some models
fit.FI <- brm(abs(FI-1) ~  cAA + cBA + cBB + cAB + sigA + sigB + phiA + phiB, data = dat, cores = 4)
fit.ND <- brm(ND ~ cAA + cBA + cBB + cAB + sigA + sigB + phiA + phiB, data = dat, cores = 4)
fit.coexist <- brm(coexist ~ cAA + cBA + cBB + cAB + sigA + sigB + phiA + phiB, data = dat, 
                   family = bernoulli, cores = 4)

# extracting coefficients and CIs
dat2 <- cbind.data.frame(estimate = c(unlist(summary(fit.FI)[["fixed"]]["Estimate"]), 
                                      unlist(summary(fit.ND)[["fixed"]]["Estimate"]), 
                                      unlist(summary(fit.coexist)[["fixed"]]["Estimate"])), 
                         metric = c(rep("FI", length(unlist(summary(fit.FI)[["fixed"]]["Estimate"]))), 
                                    rep("ND", length(unlist(summary(fit.ND)[["fixed"]]["Estimate"]))), 
                                    rep("CX", length(unlist(summary(fit.coexist)[["fixed"]]["Estimate"])))), 
                         lower = c(unlist(summary(fit.FI)[["fixed"]]["l-95% CI"]), 
                                   unlist(summary(fit.ND)[["fixed"]]["l-95% CI"]), 
                                   unlist(summary(fit.coexist)[["fixed"]]["l-95% CI"])),
                         upper = c(unlist(summary(fit.FI)[["fixed"]]["u-95% CI"]), 
                                   unlist(summary(fit.ND)[["fixed"]]["u-95% CI"]), 
                                   unlist(summary(fit.coexist)[["fixed"]]["u-95% CI"])), 
                         par = c(rownames(summary(fit.FI)[["fixed"]]),
                                 rownames(summary(fit.ND)[["fixed"]]),
                                 rownames(summary(fit.coexist)[["fixed"]])))

ggplot(data = dat2[which(dat2$par != "Intercept"),], aes(x = par, y = estimate, fill = metric, 
                                                         ymin = lower, ymax = upper)) + 
  geom_col(position = position_dodge()) + theme_classic() + 
  scale_fill_manual(name = "Metric", values = c( "black", "#F5D544", "#583B64"), 
                    labels = c("Coexistence", "Fitness Inequalities", "Niche Differences")) + 
  theme(axis.text.x = element_text(angle = 290, vjust = 0.5, hjust=0, size = 15)) + 
  scale_x_discrete(labels = c(bquote(C[AA]), bquote(C[AB]), bquote(C[BA]), bquote(C[BB]), 
                              bquote(phi[A]), bquote(phi[B]), bquote(sigma[A]), bquote(sigma[B]))) + 
  xlab("Parameter") + ylab("Effect Size") + 
  geom_errorbar(position = position_dodge(width = .9), width = .25)
```

Everything seems intuitive enough about this figure except for the fitness inequalities. It is not clear to me why increasing any given $c$ should decrease the fitness inequalities. The remainder of this document outlines my approach to rectifying this descrepency. 

### Attempt #1: Do small increases in c really decrease fitness inequalities?

First, I checked to see if this trend actually plays out with my random parameters by introducing small increases in a given $c$ value and recalculating the fitness inequalities. I begin by taking an arbitrary set of parameter values from my simulated dataset and calculate a baseline fitness inequality and niche difference value. 

```{r}
i = 22 # arbitrary set of parameters
alphaAA <- (cAAs[i]+0)+sigAs[i]*phiAs[i]
alphaBA <- (cBAs[i]+0)+sigBs[i]*phiAs[i]
alphaBB <- (cBBs[i]+0)+sigBs[i]*phiBs[i]
alphaAB <- (cABs[i]+0)+sigAs[i]*phiBs[i]

fi <- abs(ifelse(sqrt((alphaAA*alphaAB)/(alphaBB*alphaBA)) > sqrt((alphaBB*alphaBA)/(alphaAA*alphaAB)),
       sqrt((alphaAA*alphaAB)/(alphaBB*alphaBA)), sqrt((alphaBB*alphaBA)/(alphaAA*alphaAB))) - 1)
nd <- 1 - sqrt((alphaAB*alphaBA)/(alphaAA*alphaBB))
print(paste("fi = ", round(fi, 3), " and nd = ", round(nd, 3), sep = ""))
```

I then increase one of the intraspecific c's. I intuit that this should increase niche differences and fitness inequalities. However, my model predicts that this should increase niche differences and decrease fitness inequalities. 

```{r}
alphaAA <- (cAAs[i]-0.5)+sigAs[i]*phiAs[i]
alphaBA <- (cBAs[i]+0)+sigBs[i]*phiAs[i]
alphaBB <- (cBBs[i]+0)+sigBs[i]*phiBs[i]
alphaAB <- (cABs[i]+0)+sigAs[i]*phiBs[i]

fi <- abs(ifelse(sqrt((alphaAA*alphaAB)/(alphaBB*alphaBA)) > sqrt((alphaBB*alphaBA)/(alphaAA*alphaAB)),
       sqrt((alphaAA*alphaAB)/(alphaBB*alphaBA)), sqrt((alphaBB*alphaBA)/(alphaAA*alphaAB))) - 1)
nd <- 1 - sqrt((alphaAB*alphaBA)/(alphaAA*alphaBB))
print(paste("fi = ", round(fi, 3), " and nd = ", round(nd, 3), sep = ""))
```
